Value-Suppressing Uncertainty Maps

### 2018-01-03 paper/main.tex
Fleshed out discussion section.

In some cases, these assumptions are violated. For instance, an analyst might be interested in ``long tail risks'' or other ``black swan events,'' where the impact of a value, no matter how uncertain, must be considered and planned for~\cite{taleb2011black}. Other analysis tasks (such as filtering out outliers), require increased, rather than decreased, discriminability when uncertainty is high. However, for many information fusion tasks, the assumption is that uncertainty is related to data quality, or the uncertainty of data values~\cite{riveiro2007evaluation}.

@@

Using a public data set from Wikipedia~\cite{wiki}, we collected the statewide polls that were conducted closest to the 2016 U.S. presidential election, we created a choropleth map showing which candidate was leading in the polls. We calculated uncertainty as the size of the lead in terms of margin of error (lead/margin). The AP style guide for reporters suggests that a candidate leading by more than two margins of error is said to be ``significantly ahead'' in the polls~\cite{french1988associated}. As the lead becomes closer to the margin of error, there is more doubt about the electoral outcome.

By providing more color categories when the margin of error was relatively small, the VSUP map shows that, in many of the states leaning towards Hillary Clinton, her lead was quite small. The VSUP map, by limiting inferences about which candidate was leading when the uncertainty is high, also prevents the viewer from making strong claims about the outcome of the election in highly volatile states, even if a state is leaning one way or the other. Given the relative volatility in polling data, this caution is warranted. An additional benefit to the VSUP map is that, as the diverging blue-red color map passes through a central whitepoint, it can be difficult to distinguish between uncertainty levels (which also interpolates to white) in the 2D map. The VSUP's tree-like structure separates these problematic colors: the closest colors in the 2D scale are 2.9 units apart in CIELab space, but the closest colors are 6.2 units apart in the VSUP scale.

@@

We had two hypotheses, stemming from our belief that VSUPs promote better \emph{integration} between uncertainty and value information, and encourage \emph{caution} by highlighting the ambiguity introduced by uncertain data. In particular:
\begin{enumerate}
	\item Participants would avoid targets with \textbf{high uncertainty} when using a VSUP.
	\item This would result in a tradeoff where they would also choose targets with lower expected value \textbf{higher danger} when using a VSUP.
\end{enumerate}

@@

These two results, taken together, suggest that VSUPs led participants to moderate their guesses: avoiding highly uncertain regions, but willing to ``gamble'' more on regions of middling certainty. In our design, where these gambles were usually poor (due to the predominance of ``bad'' cells), this moderation resulted in guesses with higher average danger. However, when expected value and uncertainty are orthogonal, this moderation can be beneficial, or at least not harmful. Indeed, for guesses with middling and low uncertainty (quartiles 1-3), where this orthogonality was present, there was no significant difference ($t=1.56$, $p=0.11$) in average danger between VSUP ($M=0.30$, $SD=0.24$) and traditional ($M=0.29$, $SD=0.24$) conditions.

@@

However, our results indicate that juxtaposed visualizations have a substantial drawback for information fusion tasks.

@@

However, for bivariate color maps, where often the color properties used to encode value and uncertainty are not perceptually separable, error in decoding values can be quite high.  In our identification experiment, this perceptual error was high. Continuous maps performed the worst of all of our other bivariate map types (47\% accuracy vs. 63\% accuracy for discrete maps). We believe that the main driver for this poor performance

\subsection{Risk Aversion vs. Risk Seeking}

It is often difficult for the general audience to successfully make use of uncertainty information.  In order to minimize the complexity of uncertainty, uncertainty information is often visualized in a binary way, where values either meet some explicit uncertainty standard, or do not. These binary encodings for uncertain information, such as error bars for estimates, or ``cones of uncertainty'' for hurricane paths, result in binary patterns of decision-making that do not accurately reflect the underlying uncertainty in the data~\cite{cox2013visualizing,correll2014error}.

We designed VSUPs to promote risk-averse decision-making. In the limit of uncertainty VSUPs have only one output color, and so explicitly prevent any decoding of value. This narrowing of options is reflected not only in the quantization scheme, but also in the shape of the legend, narrowing as the uncertainty increases. This is despite the fact that the continuous maps were constructed via interpolations in CIELAB colorspace, which was designed for perceptual uniformity~\cite{mahy1994evaluation}. The integrality of the value and uncertainty channels 
%\sizeFig

>>>>

In some cases, these assumptions are violated. For instance, an analyst might be interested in ``long tail risks'' or other ``black swan events,'' where the impact of a value, no matter how uncertain, must be considered and planned for~\cite{taleb2011black}. Other analysis tasks (such as filtering out outliers), require increased, rather than decreased, discriminability when uncertainty is high. However, for many information fusion tasks, the assumption is that uncertainty is related to data quality, or the variability of data values~\cite{riveiro2007evaluation}.

@@

Using a public data set from Wikipedia~\cite{wiki} of the statewide polls that were conducted closest to the 2016 U.S. presidential election, we created a choropleth map showing which candidate was leading in the polls. We calculated uncertainty as the size of the lead in terms of margin of error (lead/margin). The AP style guide for reporters suggests that a candidate leading by more than two margins of error is said to be ``significantly ahead'' in the polls~\cite{french1988associated}. As the lead becomes closer to the margin of error, there is more doubt about the electoral outcome.

By providing more color categories when the margin of error was relatively small, the VSUP map shows that, in many of the states leaning towards Hillary Clinton, her lead was quite small. The VSUP map, by limiting inferences about which candidate was leading when the uncertainty is high, also prevents the viewer from making strong claims about the outcome of the election in highly volatile states, even if a state is leaning one way or the other. Given the relative volatility in polling data, this caution is warranted.

@@

We had two hypotheses, stemming from our belief that VSUPs promote better \emph{integration} between uncertainty and value information, and encourage \emph{caution} by highlighting the ambiguity introduced by uncertain data. In particular:
\begin{enumerate}
	\item Participants would avoid targets with \textbf{high uncertainty} when using a VSUP.
	\item This would result in a tradeoff where they would also choose targets with \textbf{higher danger} when using a VSUP.
\end{enumerate}

@@

These two results, taken together, suggest that VSUPs led participants to moderate their guesses: avoiding highly uncertain regions, but willing to ``gamble'' more on regions of middling certainty (as in \figref{fig:taskTwoHeatmap}). In our design, where these gambles were usually poor (due to the predominance of ``bad'' cells), this moderation resulted in guesses with higher average danger. However, when expected value and uncertainty are orthogonal, this moderation can be beneficial, or at least not harmful. Indeed, for guesses with middling and low uncertainty (quartiles 1-3), where this orthogonality was present, there was no significant difference ($t=1.56$, $p=0.11$) in average danger between VSUP ($M=0.30$, $SD=0.24$) and traditional ($M=0.29$, $SD=0.24$) conditions.

@@

However, our results indicate that juxtaposed visualizations have a substantial drawback for information fusion tasks. Integrating information from juxtaposed heatmaps is error-prone, as it requires searching for connections in two separate charts. In our experiments, the error introduced by juxtaposition outweighed the benefits of having simpler univariate scales rather than bivariate ones (accuracy of 58\% vs. 51\% for superimposed and juxtaposed charts, respectively). In our experimental data, where value and uncertainty were not well-correlated, there were no reliable visual landmarks in common amongst the value and uncertainty charts, which may have made it difficult for participants to perform quick correspondence between value in one chart, and uncertainty in the other. Given this error-prone information fusion, we recommend either the use of superposition or explicit encoding for information fusion tasks involving uncertainty. At the very least, designers should use interaction (such as highlighting across charts) to make comparison in dense juxtaposed charts easier.

@@

However, for bivariate color maps, where often the color channels for each variable are not perceptually separable, it can be difficult to decode the color into its component parts. In our identification experiment this perceptual decoding error was high. Continuous maps performed the worst of all of our other bivariate map types (47\% accuracy vs. 63\% accuracy for discrete maps). Echoing the results of Padilla et al.~\cite{padilla2017evaluating}, we hold that the error introduced by quantization, which is known and bounded, has performance benefits even for continuous data, if the perceptual error is sufficiently high. Bivariate maps, where the integrality of channels raise this perceptual error, are a prime example of such a scenario. We advise designers to avoid continuous bivariate color scales when possible, and instead employ discrete scales with a relatively small number of categories.

\subsection{Risk Averse vs. Risk Seeking Decision-Making}

It is often difficult for the general audience to successfully make use of uncertainty information.  In order to minimize the complexity of uncertainty, uncertain information is often visualized in a binary way, where values either meet some explicit uncertainty standard, or do not. These binary encodings for uncertain information, such as error bars for estimates, or ``cones of uncertainty'' for hurricane paths, result in binary patterns of decision-making that do not accurately reflect the underlying uncertainty in the data~\cite{cox2013visualizing,correll2014error}. On the other hand, presenting more detailed uncertainty information may overwhelm or confuse the viewer~\cite{greis2017designing}.

VSUPs are in a middle ground between sparse but simple, and dense but complex presentations of uncertainty. At high levels of uncertainty, VSUPs have only one output color, and so explicitly prevent any decoding of value. This narrowing of options is reflected not only in the quantization scheme, but also in the shape of the legend, narrowing as the uncertainty increases. We selected this design to promote \emph{risk-averse} decision-making. That is, people should be hesitant to make strong predictions or decisions about values when uncertainty is high. Our prediction experiment, drawing from Kahneman \& Tversky's prospect theory~\cite{kahneman1979prospect}, was further designed to elicit avoidance of uncertain data, by framing the predictions in terms of minimizing losses, rather than maximizing gains. This framing was reflected in participants mostly making safe and certain guesses (low danger, high certainty). Where the differences in decision-making occurred were where the participants had exhausted the set of safe and certain locations, and were forced to value either certainty or expected value. A risk seeker would choose uncertain locations if the reward were high enough, whereas a risk-avoider would select locations with less uncertainty, even with a lower payoff. Participants using VSUPs were likelier to make these risk-avoiding decisions than when they were using traditional bivariate maps. Beyond VSUPs, this result indicates that the designers of visualizations can have fine-grained control over the types of decision-making their visualizations promote. Visualization designers should consider the not just fidelity to the underlying dataset, but also the utility of the decisions their designs support.
%\sizeFig
